Cycle Route Uptake and Scenario Estimation (CRUSE): an open and nationally scalable approach for strategic cycle network planning tool development \\\large{A case study of Ireland}

### 2024-03-22 paper.qmd
Refactor intro

High energy transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a cause of disease due to airborne and noise pollutants.
Conversely, evidence-based and effective transport *policies* have great potential to decarbonize economies, improve public health, and save lives.

Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport.
Nearly half of these emissions (around 10% of global emissions) are from passenger cars [@jaramillo2022].
The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
Services that are only accessible by car lock-in car dependency [@gray2001; @shergold2012; @motte-baumvol2010].

Recognizing the growing evidence of such impacts of poorly designed and performing transport systems, governments in many countries have set targets and taken actions.
In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
The ASI framework highlights the importance of demand reduction (*avoid*ing unnecessary trips), in addition to mode *shift* to sustainable modes and *improvement* of existing energy converters, in that order.

@@

Another motivation for cycle network planning is the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
Better data to inform road safety policies and interventions is a motivation for the CRUSE Tool outlined in this paper, and modeling active travel more broadly.

In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030 [@national2021].
Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities.
A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
The main frameworks underpinning these efforts are the Climate Action Plan [@climate2022a], the National Development Plan, and the National Roads 2040 strategy [@national2023].
<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016 "83% of respondents supported enhanced walking and cycling facilities" [@sustaina2019].

At the regional level, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
7 in 10 trips in Ireland are by car [@national] .
Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].

Further evidence of the importance of cycling in Ireland *already* is provided by the National Strategic Objective (NSO) from the National Development Plan, with €8.6 billion allocated to sustainable transport infrastructure including public transport and active travel interventions.
Cycle infrastructure will be developed in synchrony with the BusConnects project, an entire redesign of the bus network in Dublin and Cork.

It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.

A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure, as shown in @fig-dublin.
To this end, TII extensively reviewed activity on Irish Roads and how the road network was being used by commuters, cyclists, heavy goods vehicles and other road users across the Irish transport network.
The aim was for the tool to provide strong, national, systematic but locally-specific evidence to monitor cycling friendliness and safety in Ireland, to ensure strategic alignment with national, regional and local policies.
The open source and publicly available nature of the tool ensures more inclusive and evidence-based conversations around cycle network planning between all stakeholders in the transport planning process.

TII also requires evidence to support monitoring of growth in cycling, upgrades of existing cycling infrastructure, business case development and appraisal, and exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].

::: {#fig-dublin}
<!-- ![](images/paste-1.png) -->
![](images/paste-8.png)
<!-- ![](images/paste-3.png) -->
![](images/paste-9.png)

@@

As the benefits of strategic planning for cycling become more apparent [@scappini2022] we expect the demand for tools such as CRUSE to grow.

>>>>

Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
Cars are a major and growing source of greenhouse gas emissions causing climate change [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018] , and a cause of disease due to physical inactivity and air, noise and (from tyres) microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024].
Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport.
Nearly half of these emissions (around 10% of global emissions) are from passenger cars [@jaramillo2022].
The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
Services that are only accessible by car lock-in car dependency [@gray2001; @shergold2012; @motte-baumvol2010].

Growing evidence of the negative impacts of car-dependent transport systems has led governments in many countries to set targets and taken actions.
In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
The ASI framework highlights the importance of demand reduction (*avoid*ing unnecessary trips), in addition to mode *shift* to sustainable modes and *improvement* of existing energy converters, in that order.

@@

Another motivation for cycle network planning at the European level is the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
Better data to inform road safety policies and interventions is a motivation for better estimates of 'baseline' levels of physical activity at high levels of geographic resolution [@tait2023].

National governments are increasingly acting on the evidence.
In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030 [@national2021].
Doing so while simultaneously enabling rapid uptake of active modes will require key travel corridors to be identified 'cycle proofed'.
Cycling in Ireland represents only 3% of total modal share as of the 2016 Census, but accounts for 20% of serious injuries and 7% of all fatalities, driving the development of cycling policy and infrastructure plans in Ireland.
The main frameworks underpinning these efforts are the Climate Action Plan [@climate2022a], the National Development Plan, and the National Roads 2040 strategy [@national2023].
<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016 "83% of respondents supported enhanced walking and cycling facilities" [@sustaina2019].

At the regional level within Ireland, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings of high support for and capability for cycling: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
7 in 10 trips in Ireland are by car [@national] .
Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].

Further evidence of the importance of cycling in Ireland is provided by the National Strategic Objective (NSO) from the National Development Plan, which allocates €8.6 billion to sustainable transport infrastructure including public transport and active travel interventions.
Cycle infrastructure will be developed in synchrony with the BusConnects project, an entire redesign of the bus network in Dublin and Cork.
It was in this context that Transport Infrastructure Ireland (TII) commissioned the research reported in this paper, which led to the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
Building on previous work, including the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.

Stakeholder consultation by TII emphasized the need for a strong, national, systematic but locally-specific evidence base on cycle networks in both urban and rural Ireland.
The evidence base had to scale nationally to support strategic alignment with national, regional and local policies, but also be useful for local network planning.
TII also requires

::: {#fig-dublin}
<!-- ![](images/paste-1.png) -->

![](images/paste-8.png) <!-- ![](images/paste-3.png) -->
![](images/paste-9.png)

@@

As the benefits of strategic planning for cycling become more apparent [@scappini2022] we expect the demand for tools such as CRUSE to grow.
However, the approach is not without limitations and does not meet all requirements in Ireland or indeed any country where the methods are applied.

A limitation of the approach as implemented in the CRUSE tool is that the results are 'static', reducing its ability to support monitoring of growth in cycling, upgrades of existing cycling infrastructure and dynamic (year-by-year) exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].

### 2024-03-22 paper.tex
Refactor intro

High energy transport systems are a major contributor to climate change,
a leading cause of premature death and injury due to road traffic
collisions, and a cause of disease due to airborne and noise pollutants.
Conversely, evidence-based and effective transport \emph{policies} have
great potential to decarbonize economies, improve public health, and
save lives.

>>>>

Transport systems dominated by heavy and powerful private cars are
inefficient, dangerous, and unhealthy. Cars are a major and growing
source of greenhouse gas emissions causing climate change
\citep{winkler2023}, a leading cause of premature death and injury due
to road traffic collisions \citep{globals2018} , and a cause of disease
due to physical inactivity and air, noise and (from tyres) microplastic
pollution \citep{mattsson2023} \citep{welch2023} \citep{cavallaro2024}.
